---
# IMPORTANT: Change settings here, but DO NOT change the spacing.
# Remove comments and add values where applicable.
# The descriptions below should be self-explanatory

title: "Settling the Debate: The GOAT of tennis"
#subtitle: "This will appear as Right Header"

documentclass: "elsarticle"

# --------- Thesis title (Optional - set to FALSE by default).
# You can move the details below around as you please.
Thesis_FP: FALSE
# Entry1: "An unbelievable study with a title spanning multiple lines."
# Entry2: "\\textbf{Some Guy}" # textbf for bold
# Entry3: "A thesis submitted toward the degree of Doctor of Philosophy"
# Uni_Logo: Tex/Logo.png # Place a logo in the indicated location (from your root, e.g. defaults to ~/Tex/Logo.png) and uncomment this line. Leave uncommented for no image
# Logo_width: 0.3 # If using a logo - use this to set width (size) of image
# Entry4: "Under the supervision of: \\vfill Prof. Joe Smith and Dr. Frank Smith"
# Entry5: "Stellenbosch University"
# Entry6: April 2020
# Entry7:
# Entry8:

# --------- Front Page
# Comment: ----- Follow this pattern for up to 5 authors
AddTitle: TRUE # Use FALSE when submitting to peer reviewed platform. This will remove author names.
Author1: "Grace Grant - 21653488"  # First Author - note the thanks message displayed as an italic footnote of first page.
#Ref1: "Some Institution, Cape Town, South Africa" # First Author's Affiliation
#Email1: "nfkatzke\\@gmail.com" # First Author's Email address


# Comment out below to remove both. JEL Codes only given if keywords also given.
#keywords: "Multivariate GARCH \\sep Kalman Filter \\sep Copula" # Use \\sep to separate
#JELCodes: "L250 \\sep L100"

# ----- Manage headers and footers:
#BottomLFooter: $Title$
#BottomCFooter:
#TopLHeader: \leftmark # Adds section name at topleft. Remove comment to add it.
BottomRFooter: "\\footnotesize Page \\thepage" # Add a '#' before this line to remove footer.
addtoprule: TRUE
addfootrule: TRUE               # Use if footers added. Add '#' to remove line.

# --------- page margins:
margin: 2.3 # Sides
bottom: 2 # bottom
top: 2.5 # Top
HardSet_layout: TRUE # Hard-set the spacing of words in your document. This will stop LaTeX squashing text to fit on pages, e.g.
# This is done by hard-setting the spacing dimensions. Set to FALSE if you want LaTeX to optimize this for your paper.

# --------- Line numbers
linenumbers: FALSE # Used when submitting to journal

# ---------- References settings:
# You can download cls format here: https://www.zotero.org/ - simply search for your institution. You can also edit and save cls formats here: https://editor.citationstyles.org/about/
# Hit download, store it in Tex/ folder, and change reference below - easy.
bibliography: Tex/ref.bib       # Do not edit: Keep this naming convention and location.
csl: Tex/harvard-stellenbosch-university.csl # referencing format used.
# By default, the bibliography only displays the cited references. If you want to change this, you can comment out one of the following:
#nocite: '@*' # Add all items in bibliography, whether cited or not
# nocite: |  # add specific references that aren't cited
#  @grinold2000
#  @Someoneelse2010

# ---------- General:
RemovePreprintSubmittedTo: TRUE  # Removes the 'preprint submitted to...' at bottom of titlepage
Journal: "Journal of Finance"   # Journal that the paper will be submitting to, if RemovePreprintSubmittedTo is set to TRUE.
toc: FALSE                       # Add a table of contents
numbersections: TRUE             # Should sections (and thus figures and tables) be numbered?
fontsize: 11pt                  # Set fontsize
linestretch: 1.2                # Set distance between lines.
link-citations: TRUE            # This creates dynamic links to the papers in reference list.

### Adding additional latex packages:
# header-includes:
#    - \usepackage{colortbl} # Add additional packages here.

output:
  pdf_document:
    keep_tex: TRUE
    template: Tex/TexDefault.txt
    fig_width: 3.5 # Adjust default figure sizes. This can also be done in the chunks of the text.
    fig_height: 3.5

---

<!-- First: Set your default preferences for chunk options: -->

<!-- If you want a chunk's code to be printed, set echo = TRUE. message = FALSE stops R printing ugly package loading details in your final paper too. I also suggest setting warning = FALSE and checking for warnings in R, else you might find ugly warnings in your paper. -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, fig.width = 6, fig.height = 5, fig.pos="H", fig.pos = 'H')
# Note: Include = FALSE implies the code is executed, but not printed in your pdf.
# warning and message = FALSE implies ugly messages and warnings are removed from your pdf.
# These should be picked up when you execute the command chunks (code sections below) in your rmd, not printed in your paper!

# Lets load in example data, and see how this can be stored and later called from your 'data' folder.
if(!require("tidyverse")) install.packages("tidyverse")
library(tidyverse)
library(lubridate)
library(RColorBrewer)
library(xtable)
library(caret)
library(ranger)
library(gridExtra)
list.files('code/', full.names = T, recursive = T) %>% .[grepl('.R', .)] %>% as.list() %>% walk(~source(.))

# Notice that as you are working in a .Rproj file (I am assuming you are) - the relative paths of your directories start at your specified root.
# This means that when working in a .Rproj file, you never need to use getwd() - it is assumed as your base root automatically.
```

```{r load data}
Lst_matches <- 
    list.files("data/Tennis/", full.names = T) %>% 
    .[grepl("atp_matches", .)] %>% 
    .[!grepl("doubles|qual|amateur", .)] %>% as.list()

playerinfo <-
        read_csv("data/Tennis/atp_players.csv") %>% select(player_id, hand, height)

# I want to only look at Grandslams, Masters and Finals - use Tournament_Inputs to filter.
suppressWarnings(result_Large_events <- Data_Prep_Function(Lst_matches, Tournament_Inputs = c("G", "M", "F")))

Lst_rankings <- list.files("data/Tennis/", full.names = T) %>% 
    .[grepl("atp_rankings", .)] %>% as.list()
rankings <- Ranking_Prep_Function(Lst_rankings) %>% group_by(year = lubridate::year(date)) %>% filter(date == max(date))
```



<!-- ############################## -->
<!-- # Start Writing here: -->
<!-- ############################## -->

# Introduction \label{Introduction}
The Open Era of tennis has seen the world's greatest players and some of the most riveting and exceptional performances in big events. Players such as Rod Laver, John McEnroe, Ivan Lendl and Pete Sampras made a name for themselves over the course of their careers but it is arguably the Big 3 - Roger Federer, Rafael Nadal and Novak Djokovic - who have captured the world's attention and have elevated the sport beyond anything seen before. The following machine learning project thus aims to predict which of these men should be considered the GOAT - the greatest of all time. I make use of a random forest model applied to a dataset containing all the matches in the main ATP tour events from 2003, when Federer won his first Grand Slam. However, I first provide descriptive statistics to better understand the nature of the data as well as to gain insight into the Big 3's performance.

# Descriptive statistics
Number of Grand Slams is the most commonly used metric in the debate of the best player. The following graph highlights this, showing the top 10 Grand Slam winners from 1968 to 2022.

```{r GS winners}
finals <- result_Large_events %>% filter(round == "F")
bar_top <- top_GS(finals)
bar_top
```

This graph illustrates the dominance of the Big 3 in Grand Slam wins, with Nadal holding the most titles at 22 as of 2022. The following table further supports the Big 3's supremacy, showing the winners of each Grand Slam from 2003 to 2019. 
```{r results='asis'}
merged <- GS_winners_table(finals)
GS_table <- options(xtable.comment = FALSE)
GS_table <- xtable(merged, caption = "Grand Slam Winners Since 2003")
GS_table
```

This is evidence of the extent to which Nadal, Federer and Djokovic have dominated the Grand Slam circuit. Since 2003, when Federer won his first Grand Slam, the Big 3 have won 55 out of the 68 tournaments in this period, with only only Andy Murray and Stan Wawrinka winning more than one each of all the other players. There is therefore no doubt that these top 3 players will rival each other as being the greatest of all time. 
Beyond showing Grand Slam wins, the data also provides additional information on the statistics of each match played relating to length of the game, serving and break points statistics and the surface of the tournament among others. The following tables and figures illustrate how each of the Big 3's wins are broken down and relate to some of these variables.

```{r more descriptive}
Result <- 
    c("Rafael Nadal", "Roger Federer", "Novak Djokovic") %>% as.list() %>% 
    map_df(~Winsummary_Function_Compiler(result_Large_events,
                                 Player_Name = .,
                                 playerinfo, 
                                 Top_N = 5,
                                 Silent = F)) %>% 
    filter(!grepl("Carpet", Type))

    
 filter <- Result %>% filter(grepl("Clay|Grass|Hard", Type)) %>% select(Player, Type, N, Win_Percentage)
    
    df <- filter %>% mutate(Matches_Won = Win_Percentage*N) %>% select(Player, Type, Matches_Won) %>% 
        group_by(Player) %>% summarise(fraction=Matches_Won/sum(Matches_Won)) 
    Type <- filter %>% mutate(Matches_Won = Win_Percentage*N) %>% select(Player, Type, Matches_Won) %>% 
        group_by(Player) 
    surface_merged <- cbind(df, Type) %>% select(-Player...3, -Matches_Won)
    
    
plot1 <- surface_merged %>% filter(Player...1=="Rafael Nadal") %>% 
    ggplot(aes(x="",y=fraction, fill=Type))+
    geom_bar(stat="identity",width=1)+
    coord_polar(theta="y")+
    labs(title = "Rafael Nadal", x = NULL, y = NULL)+
    theme(legend.position = "none", axis.text.x = element_blank(), axis.text.y = element_blank())
plot2 <- surface_merged %>% filter(Player...1=="Roger Federer") %>% 
    ggplot(aes(x="",y=fraction, fill=Type))+
    geom_bar(stat="identity",width=1)+
    coord_polar(theta="y")+
    labs(title = "Roger Federer", x = NULL, y = NULL)+
    theme(legend.position = "none", axis.text.x = element_blank(), axis.text.y = element_blank())
plot3 <- surface_merged %>% filter(Player...1=="Novak Djokovic") %>% 
    ggplot(aes(x="",y=fraction, fill=Type))+
    geom_bar(stat="identity",width=1)+
    coord_polar(theta="y")+
    labs(title = "Novak Djokovic", x = NULL, y = NULL)+
    theme(legend.position = "none", axis.text.x = element_blank(), axis.text.y = element_blank())
    

grid.arrange(plot1, plot2, plot3,nrow=1)

```

These graphs, referencing surfaces by their colour, show each of Nadal, Federer and Djokovic's win percentages on each surface. This indicates that Nadal has a much higher win record on clay than the other two, confirming his status as the King of Clay. All players have the highest win records on hard court, which may partly link to this being the surface with the most number of matches, but Federer and Djokovic have a more even split across surfaces, indicating their versatility.

```{r grouped bar}
bar_percent <- percentage_bar(Result)
bar_percent
```

The grouped chart sheds further light into each of the Big 3's statistics against top 5 players, averaging across matches. Djokovic has the highest win percentage, Nadal the highest breakpoints saved and Federer the highest ace percentage. Federer is known to have a strong serve that is difficult to read so it is understandable that his ace percentage is higher than the others, while Nadal is known to fight back when he is at a disadvantage, hence the high breakpoints saved. However, I would argue that win percentage is the most important statistic to consider because this relates directly to number of wins and titles. This graph shows that Djokovic has the best record against top 5 players in the main ATP events, winning approximately 62 percent of his matches. 

```{r results='asis'}
time <- Result %>% filter(grepl("All", Type)) %>% select(Player, TTaken)
time$TTaken <- round(time$TTaken)
time_table <- xtable(time, caption = "Average Time Taken")
time_table
```

This final table provides an overview of the Big 3's average match times across all games. Roger Federer is shown to take the least amount of time to finish matches, averaging at 109 minutes or approximately 1 hour and 45 minutes. This is quite significantly different from Nadal and Djokovic, suggesting that Federer prefers a shorter game format. This may link to his playing style which involves big serves and net play which generally induces shorter matches due to less rallies. 
